---
title: "spei_gis"
author: "Mitchell Stares & Eli Bendall"
date: "12/03/2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

## Libraries

### NOTE this is modified SPEI calculation, V2 - 12/03/2019

```{r}
library(raster) # Main package used for everything raster
library(tidyverse) # Normally used for data manipulations
library(rgdal) # Used mostly to read in the NSW polygon
library(sp) # Not currently needed but a useful package for vector data
library(sf) # Not currently needed but a quicker version of SP that has limited compatibility with most commands
library(RNetCDF) # NetCDF4 package
library(ncdf4) # NetCDF4 package
library(SPEI) # Used for the SPEI calculation in the second loop
library(zoo) #Used for timescale (yearmon()) stuff
library(nnet)
library(stats)
library(fasterize)
```

## PMFAO Loop  (have modified, EB, function is now set to "sum") 
This section creates the PMFAO (Precip - ET) raster for each year taking from the input folders (rain and fao). The loop also takes a NSW layer input (taken from Firelab Sharedrive) in order to crop the raster stacks to reduce processing time. Without the crop, the processing was taking along time. 

The loop:
The loop takes the list of files from the Rain folder using each entry in the list as the iterative variable. 
Each entry in the list is read in as a raster stack and then immediately cropped to the NSW extent. This is repeated for the FAO folder, which also uses the above function to calculate the monthly average ET from the daily values.  
These are then subtracted from each other to create the PMFAO file, which is written to a raster using the string split function to create a unique name for each layer.  

Strsplit() - split the individual character string by a specific character. In these examples the "/" is used followed by the "_". The second strsplit is used to split the final string from the first strsplit into two so the individual year can be obtained.



```{R}

P.nc.lst<-list.files(path="C:/Users/e_ben/OneDrive - University of Wollongong/PhD/R/SPEI 3/RAIN", pattern="rai",full.names = TRUE, recursive=TRUE, include.dirs = TRUE)
FAO.nc.lst<-list.files(path="C:/Users/e_ben/OneDrive - University of Wollongong/PhD/R/SPEI 3/FAO", pattern="sho",full.names = TRUE, recursive=TRUE, include.dirs = TRUE)
bioregions <- readOGR(dsn="C:/Users/e_ben/OneDrive - University of Wollongong/PhD/R/SPEI 3/BOUNDARIES", layer="IBRA_studyregions_ELI_2_rpj.")

# This took about 40 minutes on Mitch's computer

for(d in 1:length(P.nc.lst))
{
  P.tmp.r<-stack(P.nc.lst[d])
  P.tmp.crop <- crop(P.tmp.r, bioregions)
  FAO.tmp.r<-stack(FAO.nc.lst[d])
  FAO.tmp.crop <- crop(FAO.tmp.r, bioregions)
  indicies <- format(as.Date(names(FAO.tmp.crop), format = "X%Y.%m.%d"), format = "%m")
  indicies <- as.numeric(indicies)
  FAO.mnth.r <- stackApply(FAO.tmp.crop, indicies, fun = sum)
  PMFAO.tmp.r<-P.tmp.crop-FAO.mnth.r
  tmpfname<-strsplit(P.nc.lst[d],"/") 
  tmpfnamesplit <- strsplit(tmpfname[[1]][9], "_") #Note: Change the index here to reflect the end of strsplit for your filepath length. Eli's is [9], Mitch's is [7]
  writeRaster(PMFAO.tmp.r,paste0("C:/Users/e_ben/OneDrive - University of Wollongong/PhD/R/SPEI 3/PMFAO2/","PMFAO_",tmpfnamesplit[[1]][1],".tif"), overwrite=TRUE)
}

```


## SPEI Calculation  

```{R}

## list PMFAO files
PMFAO.lst<-list.files(path="C:/Users/e_ben/OneDrive - University of Wollongong/PhD/R/SPEI 3/PMFAO2", pattern="PMFAO",full.names = TRUE)

PMFAO.tmp.s.all <- stack(PMFAO.lst)

## Apply function - note square brackets are important 
funSPEI <- function(x, scale=12, na.rm=TRUE) {
  as.numeric((spei(x, scale=scale, na.rm=na.rm))$fitted)
}

## Calculate SPEI with function
SPEI.s.all <- calc(PMFAO.tmp.s.all, fun = funSPEI)

## Create vector of dates
dates <- format(seq(as.Date("1960-01-01"), as.Date("2017-12-01"), by="month"), '%Y%m')

## Set date names to SPEI file
SPEI.dates <- setNames(SPEI.s.all, dates)

## stack SPEI with dates
SPEI.subset <- stack(SPEI.dates)

## create vector of names for final SPEI output
SPEI.names <- names(SPEI.subset)

## write SPEI files to folder with 'suffix' as date names
writeRaster(SPEI.s.all, filename="C:/Users/e_ben/OneDrive - University of Wollongong/PhD/R/SPEI 3/SPEI_dates.tst2/SPEI", overwrite=TRUE, bylayer=TRUE, suffix=SPEI.names, format="GTiff")

```


## Subset to 6 Months if needed (dont need this currently so ignore)

```{r}#
SPEI.lst<-list.files(path="C:/Users/e_ben/OneDrive - University of Wollongong/PhD/R/SPEI 3/SPEI_dates.tst", pattern="SPEI", full.names = TRUE)

SPEI.subset <- SPEI.lst[seq(6, length(SPEI.lst), 6)]

SPEI.subset <- stack(SPEI.subset)

SPEI.names <- names(SPEI.subset)

writeRaster(SPEI.subset, filename="C:/Users/e_ben/OneDrive - University of Wollongong/PhD/R/SPEI 3/SPEI_dates.tst/SPEI", overwrite=TRUE, bylayer=TRUE, suffix=SPEI.names, format="GTiff")


```

### Drought Frequency Loop

- stack SPEI list > write function for SPEI threshold value > apply function and then calc > write raster

```{r}

SPEI.lst<-list.files(path="C:/Users/e_ben/OneDrive - University of Wollongong/PhD/R/SPEI 3/SPEI_8917", pattern="SPEI", full.names = TRUE)

r <- stack(SPEI.lst)


drought.tmp <- function(x, t = -1.5) { 
    sum(x < t, na.rm = TRUE)
  }

drought.freq.tmp <- calc(r, fun=drought.tmp)


writeRaster(drought.freq.tmp, "C:/Users/e_ben/OneDrive - University of Wollongong/PhD/R/SPEI 3/DROUGHT_FREQ/df1989-2017_-1.5.tif", format = "GTiff", overwrite=TRUE)

```

### Time-since-drought loop

Follows smilar logic to the 'drought frequency loop' above, although here we are wanting the index position of the last drought. Once obtained we can reclassify the index position to something relevant, e.g. '6 months, 12 months, 48 month etc.' to give us a 'time-since-drought' raster.

Here are the steps:

(1) Create new folder "SPEI_6m_8813" and manually copy over the files from start 1988 to end 2012, giving 50 files in the new folder.

(2) create list

(3) stack list

(4) write function - this is a modification of the 'drought freq' function above. 

```{r}
SPEI.lst<-list.files(path="C:/Users/e_ben/OneDrive - University of Wollongong/PhD/R/SPEI 3/SPEI_8917", pattern="SPEI",full.names = TRUE)

r <- stack(SPEI.lst)


last.drought <- function(x, t = -1.0) {
  imax <- max(which(x<t)) 
  ifelse(length(imax) == 0, 0, imax)
  
}




time.since.drought <- calc(r, fun=last.drought)

writeRaster(time.since.drought, "C:/Users/e_ben/OneDrive - University of Wollongong/PhD/R/SPEI 3/DROUGHT_FREQ/TSD_1989-2017_-1.0.tif", format = "GTiff", overwrite=TRUE)

```


### TSF Difference Index (SYDNEY BIOREGION)

```{r}

### read in shapefile

SYD.early.shp <- st_read("SYDregion_early_1.shp")

### replace FireYear season with actual year

SYD.early.shp$FireYear  <- str_sub(SYD.early.shp$FireYear, - 6, -3)

SYD.early.shp$FireYear <- as.numeric(SYD.early.shp$FireYear)

## read in extent object
extent.object<-raster("SYDregion_green_dbj_mosaic.tif")

#create new raster from from extent object
new.raster <- raster(extent.object)

## assign new value of zero
new.raster[new.raster] <- 0

blank.raster <- new.raster


## rasterize shp with 'max' function
SYDearlyTSF <- fasterize(SYD.early.shp, blank.raster, field="FireYear", fun='max', background=NA)


## calculate TSF

SYDearlyTSF <- 1990 - SYDearlyTSF



## repeat for next shp

### read in shapefile

SYD.late.shp <- st_read("SYDregion_late_1.shp")


### replace FireYear season with actual year

SYD.late.shp$FireYear  <- str_sub(SYD.late.shp$FireYear, - 6, -3)

SYD.late.shp$FireYear <- as.numeric(SYD.late.shp$FireYear)


## rasterize shp with 'max' function
SYDlateTSF <- fasterize(SYD.late.shp, blank.raster, field="FireYear", fun='max', background=NA)

## calculate TSF

SYDlateTSF <- 2017 - SYDlateTSF


## calculate TSF difference index

SYDregion_TSFDI <- SYDlateTSF - SYDearlyTSF


## write file


writeRaster(SYDregion_TSFDI, "SYDregion_TSFDI_14JUL20.tif", format = "GTiff", overwrite=TRUE)

```


### SEC Bioregion (NSW)

```{r}

### read in shapefile

SEC.nsw.early.shp <- st_read("SECregion_fires_early_nsw.shp")

### replace FireYear season with actual year

SEC.nsw.early.shp$FireYear  <- str_sub(SEC.nsw.early.shp$FireYear, - 6, -3)

SEC.nsw.early.shp$FireYear <- as.numeric(SEC.nsw.early.shp$FireYear)


## read in extent object
extent.object<-raster("SECregion_green_APR5th.tif")

#create new raster from from extent object
new.raster <- raster(extent.object)

## assign new value of zero
new.raster[new.raster] <- 0

blank.raster <- new.raster


## rasterize shp with 'max' function
SECnswearlyTSF <- fasterize(SEC.nsw.early.shp, blank.raster, field="FireYear", fun='max', background=NA)


## calculate TSF

SECnswearlyTSF <- 1990 - SECnswearlyTSF



## repeat for next shp

SEC.nsw.late.shp <- st_read("SECregion_fires_nsw_late.shp")

### replace FireYear season with actual year

SEC.nsw.late.shp$FireYear  <- str_sub(SEC.nsw.late.shp$FireYear, - 6, -3)

SEC.nsw.late.shp$FireYear <- as.numeric(SEC.nsw.late.shp$FireYear)




## rasterize shp with 'max' function
SECnswlateTSF <- fasterize(SEC.nsw.late.shp, blank.raster, field="FireYear", fun='max', background=NA)


## calculate TSF

SECnswlateTSF <- 2017 - SECnswlateTSF


## calculate TSF difference index

SECregion_nsw_TSFDI <- SECnswlateTSF - SECnswearlyTSF


## write file

writeRaster(SECregion_nsw_TSFDI, "SECregion_nsw_TSFDI_14JUL29.tif", format = "GTiff", overwrite=TRUE)



```


## SEC Bioregion (VIC)

```{r}

### read in shapefile

SEC.vic.early.shp <- st_read("SECregion_fires_vic_early2.shp")

SEC.vic.early.shp$SEASON2 <- as.numeric(SEC.vic.early.shp$SEASON2)



## read in extent object
extent.object<-raster("SECregion_green_APR5th.tif")

#create new raster from from extent object
new.raster <- raster(extent.object)

## assign new value of zero
new.raster[new.raster] <- 0

blank.raster <- new.raster


## rasterize shp with 'max' function
SECvicearlyTSF <- fasterize(SEC.vic.early.shp, blank.raster, field="SEASON2", fun='max', background=NA)


## calculate TSF

SECvicearlyTSF <- 1990 - SECvicearlyTSF



## repeat for next shp

SEC.vic.late.shp <- st_read("SECregion_fires_vic_late2.shp")


SEC.vic.late.shp$SEASON2 <- as.numeric(SEC.vic.late.shp$SEASON2)



## rasterize shp with 'max' function
SECviclateTSF <- fasterize(SEC.vic.late.shp, blank.raster, field="SEASON2", fun='max', background=NA)


## calculate TSF

SECviclateTSF <- 2017 - SECviclateTSF


## calculate TSF difference index

SECregion_vic_TSFDI <- SECviclateTSF - SECvicearlyTSF


## write file

writeRaster(SECregion_vic_TSFDI, "SECregion_vic_TSFDI3_14JUL2020.tif", format = "GTiff", overwrite=TRUE)


```

```{r}
SECregion_TSFDI_comb <- raster::merge(SECregion_vic_TSFDI, SECregion_nsw_TSFDI)

writeRaster(SECregion_TSFDI_comb, "SECregion_TSFDI_comb_14JUL2020.tif", format = "GTiff", overwrite=TRUE)
```

