---
title: "SYDBIO_GAMs"
author: "Eli Bendall & Micheal Bedward"
date: "02/08/2020"
output:
  word_document: default
  html_document: default
---

```{r setup, include=TRUE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(dplyr)
library(ggplot2)
library(readxl)
library(stringr)
library(tidyr)
library(stats)
library(purrr)
library(rstan)
library(brms)
library(mgcv)
library(future)

#spatial library
library(class)
library(spdep)
library(RANN)
library(sf)
library(sp)
library(lwgeom)
library(spatialEco)


#These options help Stan run faster:

rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

# Make sure the scales package is available (it should be if ggplot is installed)
requireNamespace("scales")

# Default graph theme - white background
theme_set( theme_bw() )

set.seed(42)


# Set to TRUE to force models to be refitted
REFIT_MODELS <- FALSE


# Create a folder for fitted models if one does not already exist
FittedModelsPath <- here("fitted_models_SYDBIO")
if (!dir.exists(FittedModelsPath)) {
  dir.create(FittedModelsPath)
}


# Load brms if it is installed (requires Stan software and the rstan package).
# If not installed, HAS_BRMS will be set to FALSE.
HAS_BRMS <- suppressWarnings(
  require("brms", quietly = TRUE)
)


###### Some helper functions #####

# Calculate standard page sizes
pagesize <- function(size = c("A4", "A3", "A2", "A1", "A0"), 
                     orientation = c("portrait", "landscape"),
                     units = c("cm", "mm")) {
  
  size <- match.arg(size)
  orientation <- match.arg(orientation)
  units <- match.arg(units)
  
  alpha <- 1000 * 2^(1/4)
  i <- as.integer(substr(size, 2, 2))
  long <- alpha * 2^(-i/2)
  
  page <- switch(
    orientation,
    portrait = c(width = long / sqrt(2), height = long),
    landscape = c(width = long, height = long / sqrt(2))
  )
  
  page <- round(page)
  if (units == "cm") page <- page / 10
  
  page <- c(as.list(page), units = units)
  class(page) <- "pagesize"
  
  page
}



# Save a graph to a PDF file
gg_pdf <- function(plot, filename, size = pagesize("A4", "landscape", "cm")) {
  
  if (!inherits(size, "pagesize")) stop("The size argument should be a pagesize (list) object")
  
  ggsave(
    filename, 
    plot, 
    width = size$width,
    height = size$height,
    units = size$units)
}


# Calculate highest posterior density interval for a vector of values
hpdi.vec <- function (x, prob = 0.95) {
  n <- length(x)
  if (n <= 1) stop("x must have more than 1 element")
  x <- sort(x)

  gap <- max(1, min(n - 1, round(n * prob)))
  init <- 1:(n - gap)

  inds <- which.min(x[init + gap] - x[init])

  out <- c(lower = x[inds], upper = x[inds + gap])
  out
}





```


## Read in data

```{r}

### Load stack

DSF.mask <- raster::stack("DSF.mask.SYD.2.tif")

WSF.mask <- raster::stack("WSF.mask.SYD.2.tif")

### names of .value reflect only the orignal sf object, need to change names of raster value columns for continuity

names(DSF.mask) <- c("green", "TSD", "TSFDI", "veg", "bio01", "bio12")

names(WSF.mask) <- c("green", "TSD", "TSFDI", "veg", "bio01", "bio12")

```


```{r}
### Convert stack to data frame, retaining x y

DAT.SYD.DSF.temp <- raster::as.data.frame(DSF.mask, xy = TRUE)


### filter NA

DAT.SYD.DSF.temp2 <- DAT.SYD.DSF.temp %>%
    filter(!is.na(green)) %>%
    filter(!is.na(TSD)) %>%
    filter(!is.na(TSFDI)) %>%
    filter(!is.na(bio01)) %>%
    filter(!is.na(bio12)) 

### filter TSFDI values to -10, 10

DAT.SYD.DSF.temp3 <- DAT.SYD.DSF.temp2 #%>%


### scale index

DAT.SYD.DSF.temp3$green[DAT.SYD.DSF.temp3$green > 201] <- 201

DAT.SYD.DSF.temp3$green <- (DAT.SYD.DSF.temp3$green*0.01)+(-1.01)


### Randomly sample 10000 rows

DAT.SYD.DSF.a <- as.data.frame(DAT.SYD.DSF.temp3[sample(nrow(DAT.SYD.DSF.temp3), 150000), ])

DAT.SYD.DSF.b <- as.data.frame(DAT.SYD.DSF.temp3[sample(nrow(DAT.SYD.DSF.temp3), 150000), ])

DAT.SYD.DSF.c <- as.data.frame(DAT.SYD.DSF.temp3[sample(nrow(DAT.SYD.DSF.temp3), 150000), ])


### Repeat for WSF
  
DAT.SYD.WSF.temp <- raster::as.data.frame(WSF.mask, xy = TRUE)

DAT.SYD.WSF.temp2 <- DAT.SYD.WSF.temp %>%
    filter(!is.na(green)) %>%
    filter(!is.na(TSD)) %>%
    filter(!is.na(TSFDI)) %>%
    filter(!is.na(bio01)) %>%
    filter(!is.na(bio12)) 

DAT.SYD.WSF.temp3 <- DAT.SYD.WSF.temp2 #%>%


### scale index

DAT.SYD.WSF.temp3$green[DAT.SYD.WSF.temp3$green > 201] <- 201

DAT.SYD.WSF.temp3$green <- (DAT.SYD.WSF.temp3$green*0.01)+(-1.01)


DAT.SYD.WSF.a <- as.data.frame(DAT.SYD.WSF.temp3[sample(nrow(DAT.SYD.WSF.temp3), 150000), ])

DAT.SYD.WSF.b <- as.data.frame(DAT.SYD.WSF.temp3[sample(nrow(DAT.SYD.WSF.temp3), 150000), ])

DAT.SYD.WSF.c <- as.data.frame(DAT.SYD.WSF.temp3[sample(nrow(DAT.SYD.WSF.temp3), 150000), ])

```

```{r}

DAT.SYD.DSF.a.tst <- DAT.SYD.DSF.a %>%

  mutate(TSFDI = factor(TSFDI))



ggplot(data = DAT.SYD.DSF.a.tst, aes(x = green, col = TSFDI)) +
  geom_density()
```

```{r}

DAT.SYD.DSF.b.tst <- DAT.SYD.DSF.b %>%

  mutate(TSFDI = factor(TSFDI))



ggplot(data = DAT.SYD.DSF.b.tst, aes(x = green, col = TSFDI)) +
  geom_density()
```

```{r}

DAT.SYD.DSF.c.tst <- DAT.SYD.DSF.c %>%

  mutate(TSFDI = factor(TSFDI))



ggplot(data = DAT.SYD.DSF.c.tst, aes(x = green, col = TSFDI)) +
  geom_density()
```

```{r}

DAT.SYD.WSF.a.tst <- DAT.SYD.WSF.a %>%

  mutate(TSFDI = factor(TSFDI))



ggplot(data = DAT.SYD.WSF.a.tst, aes(x = green, col = TSFDI)) +
  geom_density()
```


```{r}

DAT.SYD.WSF.b.tst <- DAT.SYD.WSF.b %>%

  mutate(TSFDI = factor(TSFDI))



ggplot(data = DAT.SYD.WSF.b.tst, aes(x = green, col = TSFDI)) +
  geom_density()
```


```{r}

DAT.SYD.WSF.c.tst <- DAT.SYD.WSF.c %>%

  mutate(TSFDI = factor(TSFDI))



ggplot(data = DAT.SYD.WSF.c.tst, aes(x = green, col = TSFDI)) +
  geom_density()
```






## Create dataframe and factors

```{r}

DAT.SYD.DSF.all <- DAT.SYD.DSF.a %>%
  dplyr::select(tsd = TSD,
         green = green,
         tsfdi = TSFDI,
         mat = bio01,
         mar = bio12,
         x = x,
         y = y)


### recalculate TSD

DAT.SYD.DSF.all$tsd <- 327 - DAT.SYD.DSF.all$tsd


DAT.SYD.WSF.all <- DAT.SYD.WSF.c %>%
  dplyr::select(tsd = TSD,
         green = green,
         tsfdi = TSFDI,
         mat = bio01,
         mar = bio12,
         x = x,
         y = y)


### recalculate TSD

DAT.SYD.WSF.all$tsd <- 327 - DAT.SYD.WSF.all$tsd

### remove clutter from workspace

rm(DAT.SYD.DSF.temp, DAT.SYD.DSF.temp2, DAT.SYD.DSF.temp3)
rm(DAT.SYD.WSF.temp, DAT.SYD.WSF.temp2, DAT.SYD.WSF.temp3, DSF.mask, WSF.mask)
  
```


## Check for missing values

```{r}
colSums( is.na(DAT.SYD.DSF.all) )
```


```{r}
colSums( is.na(DAT.SYD.WSF.all) )
```


```{r}
### Subsample to points no closer than 300m distance

library(spatialEco)
library(sp)
coordinates(DAT.SYD.DSF.all) <- ~x+y

sub.DAT.SYD.DSF <- subsample.distance(DAT.SYD.DSF.all, size = 12000, d = 0.003, echo = FALSE)  

DAT.SYD.DSF_300m <- as.data.frame(sub.DAT.SYD.DSF)

save(DAT.SYD.DSF_300m, file = "DAT.SYD.DSF_300m.2.RData")
    
```

```{r}
DAT.SYD.DSF_300m.tst <- DAT.SYD.DSF_300m %>%
  mutate(tsfdi = factor(tsfdi))

ggplot(data = DAT.SYD.DSF_300m.tst, aes(x = green, col = tsfdi)) +
  geom_density()
```



```{r}

### Repeat for WSF

coordinates(DAT.SYD.WSF.all) <- ~x+y

sub.DAT.SYD.WSF.a <- subsample.distance(DAT.SYD.WSF.all, size = 10000, d = 0.003, echo = FALSE)

sub.DAT.SYD.WSF.b <- subsample.distance(DAT.SYD.WSF.all, size = 10000, d = 0.003, echo = FALSE)

sub.DAT.SYD.WSF.c <- subsample.distance(DAT.SYD.WSF.all, size = 10000, d = 0.003, echo = FALSE)

DAT.SYD.WSF_300m <- as.data.frame(sub.DAT.SYD.WSF.a)

save(DAT.SYD.WSF_300m, file = "DAT.SYD.WSF_300m.2.RData")


```

```{r}
DAT.SYD.WSF_300m.tst <- DAT.SYD.WSF_300m %>%
  filter(tsfdi > -15) %>%
  mutate(tsfdi = factor(tsfdi))

ggplot(data = DAT.SYD.WSF_300m.tst, aes(x = green, col = tsfdi)) +
  geom_density()
```
### run model 1 for DSF

```{r}

DSF.model.SYD <- gam(green ~ te(x, y, k = 30) + s(tsd) + s(tsfdi) + s(mat) + s(mar), 
          data = DAT.SYD.DSF_300m)

save(DSF.model.SYD, file = "DSF.model.SYD.2.RData")


```

Model summary (similar to that from lme4::glmer). We only use this as a sanity check - not directly for inference.

```{r}

summary(DSF.model.SYD)

```

These values are very similar to those produced by lme4::glmer.


Graphical summaries of fitted parameters (densities and trace plots). Again these are mostly for model checking.

```{r}

plot(DSF.model.SYD, ask = FALSE)

```

```{r}
### model 1

DAT.SYD.WSF_300m <- DAT.SYD.WSF_300m %>%
  filter(tsfdi > -15)

save(DAT.SYD.WSF_300m, file = "DAT.SYD.WSF_300m.2.RData")
```
### run model WSF

```{r}

WSF.model.SYD <- gam(green ~ te(x, y, k = 30) + s(tsd) + s(tsfdi) + s(mat) + s(mar), 
          data = DAT.SYD.WSF_300m)

save(WSF.model.SYD, file = "WSF.model.SYD.2.RData")

```

Model summary (similar to that from lme4::glmer). We only use this as a sanity check - not directly for inference.



```{r}

summary(WSF.model.SYD)

```

These values are very similar to those produced by lme4::glmer.


Graphical summaries of fitted parameters (densities and trace plots). Again these are mostly for model checking.

```{r}

plot(WSF.model.SYD, ask = FALSE)

```


